Eco-evolutionary dynamics of finite populations from first principles

### 2023-02-14 ind_files/99_Appendix_B.tex
Streamline derivation of trait frequency dynamics

where we have defined the total (scaled) population size\footnote{This is $N_{K}(t)$ in the main text, but we omit the subscript here to reduce notational clutter}  $N = \sum_i x_i$ and the frequency of the $i\textsuperscript{th}$ type $p_i = f_i(x)$ and used the fact that $\frac{\partial x_j}{\partial x_k} = \delta_{jk}$. The $jk\textsuperscript{th}$ element of the Hessian is given by:

@@

If the birth and death rates take the form of \eqref{nD_functional_forms_for_replicator}, then it is easy to see using \eqref{nD_det_limit_fitess_defn} that equation \eqref{nD_for_Ito_first_term} is exactly the RHS of \eqref{nD_replicator_intermediate_1}. Thus, when the birth and death rates take the form of \eqref{nD_functional_forms_for_replicator}, then \eqref{nD_for_Ito_first_term} describes the deterministic component of the dynamics as described by the replicator-mutator equation, Price equation, etc. in the infinite population limit. These are the effects of selection and mutation at the infinite population limit. However, the finiteness of the population adds a second term to these dynamics, described by the second term that multiplies $dt$ in \eqref{nD_Ito_formula}. To calculate it, we first calculate:

@@

Thus, using \eqref{nD_jacobian_for_ito}, we see that the last term on the RHS of \eqref{nD_Ito_formula} is given by

@@

@@

N(t) dp_i &= \left(A^{-}_{i} - p_i\sum\limits_{j=1}^{m}A^{-}_{j}\right)dt - \frac{1}{K}\frac{1}{N(t)}\left(A^{+}_{i}-p_i\left(\sum\limits_{k=1}^{m} A^{+}_k\right)\right)dt\\
&+ \frac{1}{\sqrt{K}}\left[\left(A^{+}_{i}\right)^{1/2}dB^{(i)}_t - p_i\sum\limits_{j=1}^{m}\left(A^{+}_{j}\right)^{1/2}dB^{(j)}_t\right]

@@

Substituting the functional forms given by \eqref{nD_functional_forms_for_replicator} and repeating calculations for the $A^{+}_i$ terms exactly analogous to those done in going from \eqref{nD_replicator_intermediate_1} to \eqref{nD_replicator_mutator} now yields equation \eqref{nD_eqn_for_frequencies} in the main text.

>>>>

where we have defined the total (scaled) population size\footnote{This is $N_{K}(t)$ in the main text, but we omit the subscript here to reduce notational clutter} $N = \sum_i x_i$ and the frequency of the $i\textsuperscript{th}$ type $p_i = f_i(x)$ and used the fact that $\frac{\partial x_j}{\partial x_k} = \delta_{jk}$.
The $jk\textsuperscript{th}$ element of the Hessian is given by:

@@

This term describes the effects of selection and mutation at the infinite population limit. However, the finiteness of the population adds a second directional term to these dynamics, described by the second term that multiplies $dt$ in \eqref{nD_Ito_formula}. To calculate it, we first calculate:

@@

Thus, using \eqref{nD_jacobian_for_ito}, we see that the last term on the RHS of \eqref{nD_Ito_formula} is given by:

@@

\label{nD_general_stoch_freq_change}

@@

dp_i &= \underbrace{\frac{1}{N(t)}\left(A^{-}_{i} - p_i\sum\limits_{j=1}^{m}A^{-}_{j}\right)dt}_{\text{$K \to \infty$ prediction}} - \frac{1}{K}\underbrace{\frac{1}{N^2(t)}\left(A^{+}_{i}-p_i\left(\sum\limits_{k=1}^{m} A^{+}_k\right)\right)dt}_{\substack{\text{Directional finite size effects}\\\text{due to differential turnover rates}}}\\
&+ \frac{1}{\sqrt{K} N(t)}\underbrace{\left[\left(A^{+}_{i}\right)^{1/2}dB^{(i)}_t - p_i\sum\limits_{j=1}^{m}\left(A^{+}_{j}\right)^{1/2}dB^{(j)}_t\right]}_{\substack{\text{Non-directional finite size effects}\\\text{due to stochastic fluctuations}}}

@@

Plugging the functional forms of \eqref{nD_functional_forms_for_replicator} and the definitions of $w_i$ and $\tau_i$ into the definitions of $A^{-}_i$ and $A^{+}_i$, we obtain the relations
\label{nD_det_limit_fitness_and_turnover}
A^{-}_i &= x_iw_i(\mathbf{x}) + \mu Q_i(\mathbf{x})\\
A^{+}_i &= x_i\tau_i(\mathbf{x}) + \mu Q_i(\mathbf{x})
Thus, for the first term of \eqref{nD_general_stoch_freq_change}, we have
\frac{1}{N(t)}\left(A^{-}_{i} - p_i\sum\limits_{j=1}^{m}A^{-}_{j}\right) &=  \frac{1}{N(t)}\left[w_i(\mathbf{x})x_i + \mu Q_i(\mathbf{x})\right] - \frac{p_i}{N(t)}\sum\limits_{j=1}^{m}\left[w_j(\mathbf{x})x_j + \mu Q_j(\mathbf{x})\right]\\
&= w_i(\mathbf{x})p_i + \frac{\mu}{N(t)}Q_i(\mathbf{x}) - p_i\sum\limits_{j=1}^{m}\left[w_j(\mathbf{x})p_j + \frac{\mu}{N(t)}Q_j(\mathbf{x})\right]
Where we have used the definition of $p_i$ from \eqref{nD_tot_pop_and_prop_inds_defn}. Now using the definition of mean fitness from \eqref{nD_mean} and rearranging terms gives us
\frac{1}{N(t)}\left(A^{-}_{i} - p_i\sum\limits_{j=1}^{m}A^{-}_{j}\right) = (w_i(\mathbf{x}) - \overline{w})p_i + \mu\left[Q_i(\mathbf{p}) - p_i\left(\sum\limits_{j=1}^{m}Q_j(\mathbf{p})\right)\right]
where we have defined $Q_j(\mathbf{p}) = Q_j(\mathbf{x})/N(t)$. Repeating the exact same calculations for the $A^{+}_i$ terms in the second term of \eqref{nD_general_stoch_freq_change} now yields equation \eqref{nD_eqn_for_frequencies} in the main text.
